---
title: "XIWEN_RNA_SEQ"
author: "Sara Nicholson"
date: "2024-01-16"
output: html_document
---
# EdgeR #

### Initializations
```{r}
library(edgeR)
library(dplyr)
library(tibble)
```

**Read Counts File**
```{r}
# Read in your count table
x <- read.csv("/XIWEN/RNA-SEQ-FQ-2024/ANALYSIS/count_table.csv", row.names="features")
head(x)
x <- x %>% dplyr::select("S13", "S14", "S5", "S8",  "S4", "S6", "S7", "S9", "S10", "S11", "S12", "S15") 

# load biomart package
library(biomaRt)

# Create vector of ensembl gene IDs
ENSG <- as.vector(row.names(x))

# define biomart object
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
# query biomart - retrieve corresponding gene names to ENSG
results <- getBM(attributes = c("ensembl_gene_id_version", "external_gene_name", "ensembl_gene_id"),
               filters = "ensembl_gene_id_version", values = ENSG,
               mart = mart)
head(results)

# Write conversion chart to CSV
# write.csv(results, "/XIWEN/RNA-SEQ-FQ-2024/ANALYSIS/ENSG_symbol_conversions.csv")
```

```{r}
# Create Factored Groups from Column Names 
# Must be in same order as column names on count table
DataGroups <- factor(c("CTRL",  "CTRL",  "CTRL", "CTRL", "BICT", "CABO", "CABO",  "CABO", "BICT", "CABO", "BICT", "BICT"), levels = c("CTRL", "BICT", "CABO"))

Donor <- c("HD260", "MOLT4", "HD265", "HD264", "HD265", "HD260", "MOLT4", "HD265", "MOLT4", "HD264", "HD260", "HD264")

# Create DGEList for EdgeR
d <- DGEList(counts=x,group=DataGroups)
d$samples$donor <- factor(Donor)
head(d$samples)
```


### Filter Out Lowly Expressed Genes
Use the cpm function & remove features that are not expressed over 1 CPM in at least 3 (SMALLEST GROUP SIZE - 1) samples.
```{r}
# Check Prior Dimensions
dim(d) #62703

# Function to remove features not expressed over 1 CPM in at least 3 group
keep <- rowSums(cpm(d)>1) >= 3
d <- d[keep,]

# Check Dimensions after Filtering
dim(d) #15195

###################### edgeR recommended
dim(d) #62703
keep <- filterByExpr(d, group= d$samples$group)
 d <- d[keep, , keep.lib.sizes=FALSE]
 dim(d) # 16493
```
*17586 genes remain in analysis*


### Normalize for Library Size: TMM Normalization.
*'If a small proportion of highly expressed genes consume a substantial proportion of the total library size for a particular sample, this will cause the remaining genes to be under-sampled for that sample.....The calcNormFactors function normalizes the library sizes by finding a set of scaling factors for the library sizes that minimizes the log-fold changes between the samples for most genes.' -EdgeR*
```{r}
# set method to TMM or CPM normalization will be implemented
d <- calcNormFactors(d, method = "TMM")
# view Normalization Factors
d$samples
dim(d)

# use cpm() function to get normalized counts - applies normalization factors so it is TMM when in combination with above function
dtmm <- as.data.frame(cpm(d))
head(dtmm)

# Write TMM normalized Counts to Data Frame CSV for future use
# write.csv(dtmm, "/Shuai_RNA/ANALYSIS/LimmaEdgeR/tmm_norm_counts_all_samp.csv")
```

# Limma #
### Initializations
```{r}
library(limma)
library(Glimma)
library(dplyr)
library("AnnotationDbi")
library("org.Hs.eg.db")
library(biomaRt)
```

**Unsupervised Clustering of Samples**
```{r}
# equivalent to plotMDS of cpm(d, log = T) 
mds <- plotMDS(d, col = as.numeric(d$samples$donor), pch = 19)
mds

plotMDS(d, col = as.numeric(d$samples$group))
plotMDS(d, col = as.numeric(d$samples$group), gene.selection = "common", cex = 1, pch = 19)
```

```{r}
library(sva)
# Create Design Matrix (BLOCKING)
design <- model.matrix(~ d$samples$donor + d$samples$group) #  + d$samples$donor
design
```

# EDGE-R
```{r}
dge <- estimateDisp(d, design, robust = T)
fit <- glmQLFit(dge, design)

# BICT
qlfB <- glmQLFTest(fit, coef=5)
topTags(qlfB)

# CABO
qlfC <- glmQLFTest(fit, coef=6)
topTags(qlfC)

res_BICT=as.data.frame(topTags(qlfB, n=Inf))
head(res_BICT)
res_BICT$ensembl_gene_id_version <- row.names(res_BICT)
head(res_BICT)

# Get Tob Gene Tables for BICTvCTRL & CABOvCTRL DEA
res_CABO=as.data.frame(topTags(qlfC, n=Inf))
head(res_CABO)
res_CABO$ensembl_gene_id_version <- row.names(res_CABO)
head(res_CABO)

# Add annotations for genes
conversions <- read.csv("/XIWEN/RNA-SEQ-FQ-2024/ANALYSIS/ENSG_symbol_conversions.csv")
res <- left_join(res_BICT, conversions, by=c("ensembl_gene_id_version"))
res_BICT <- res %>% dplyr::select(external_gene_name, logCPM, PValue, logFC, FDR, ensembl_gene_id_version)

conversions <- read.csv("/XIWEN/RNA-SEQ-FQ-2024/ANALYSIS/ENSG_symbol_conversions.csv")
res <- left_join(res_CABO, conversions, by=c("ensembl_gene_id_version"))
res_CABO <- res %>% dplyr::select(external_gene_name, logCPM, PValue, logFC, FDR, ensembl_gene_id_version)

# write.csv(res_CABO, "/XIWEN/RNA-SEQ-FQ-2024/ANALYSIS/edgeR_results_CABOvCTRL.csv")
# write.csv(res_BICT, "/XIWEN/RNA-SEQ-FQ-2024/ANALYSIS/edgeR_results_BICTvCTRL.csv")

##### explore average log2 CPM expression
c <- d$counts
cpms <- cpm(c)
log2cpms <- log2(cpms)
avelog2cpms <- apply(log2cpms, 1, function(x) mean(x))
# write.csv(log2cpms, "/XIWEN/RNA-SEQ-FQ-2024/ANALYSIS/log2cpms.csv")
# write.csv(cpms, "/XIWEN/RNA-SEQ-FQ-2024/ANALYSIS/cpms.csv")


#########################################################
res_BICT$sig = as.factor(ifelse(res_BICT$FDR < 0.05 & abs(res_BICT$logFC)> 2,
ifelse(res_BICT$logFC > 2 ,'up','down'),'not'))
summary(res_BICT$sig)

res_BICT$sig = as.factor(
ifelse(res_BICT$FDR < 0.05, ifelse(res_BICT$logFC > 0 ,'up','down'),'not'))
summary(res_BICT$sig)

significant <- res_BICT %>% filter(sig == "up" | sig == "down")
dim(significant) # 0

# write.csv(significant, "/XIWEN/RNA-SEQ-FQ-2024/ANALYSIS/sig_genes_BICT_nologFCthresh.csv")
#####
res_CABO$sig = as.factor(ifelse(res_CABO$FDR < 0.05 & abs(res_CABO$logFC)> 2,
ifelse(res_CABO$logFC > 2 ,'up','down'),'not'))
summary(res_CABO$sig)

res_CABO$sig = as.factor(
ifelse(res_CABO$FDR < 0.05, ifelse(res_CABO$logFC > 0 ,'up','down'),'not'))
summary(res_CABO$sig)

significant <- res_CABO %>% filter(sig == "up" | sig == "down")
dim(significant) # 0

# write.csv(significant, "/XIWEN/RNA-SEQ-FQ-2024/ANALYSIS/sig_genes_CABO_nologFCthresh.csv")


# ##################################### average log 2 cpm expression
# c <- d$counts
# cpms <- cpm(c)
# log2cpms <- log2(cpms)
# log2cpms <- as.data.frame(log2cpms)
# log2cpms$rowname <- row.names(log2cpms)
# sslog2cpms <- log2cpms %>% filter(rowname %in% significant$ensembl_gene_id_version)
# sslog2cpms <- sslog2cpms %>% dplyr::select(hsc18, hsc20, hsc30, hsc35, hsc36, hsc3, hsc4, hsc23, hsc25, hsc33)
# avelog2cpms <- apply(as.matrix(sslog2cpms), 1, function(x) mean(x))
# 
# ####################### make CPM and count Tables
# # CPM
# cpms <- as.data.frame(cpms)
# cpms$rowname <- row.names(cpms)
# cpms <- cpms %>% filter(rowname %in% significant$ensembl_gene_id_version)
# cpms <- left_join(cpms, conversions, by=c("rowname" = "ensembl_gene_id_version"))
# 
# na_idx <- which(is.na(cpms$external_gene_name))
# cpms$external_gene_name[na_idx] <- cpms$rowname[na_idx]
# em_idx <- (which(cpms$external_gene_name == ""))
# cpms$external_gene_name[em_idx] <- cpms$rowname[em_idx]
# 
# row.names(cpms) <- cpms$external_gene_name
# cpms <- cpms %>% dplyr::select(hsc18, hsc20, hsc30, hsc35, hsc36, hsc3, hsc4, hsc23, hsc25, hsc33)
# write.csv(cpms, "./significant_cpms_edgeR.csv")
# 
# # Counts
# counts <- as.data.frame(c)
# counts$rowname <- row.names(counts)
# counts <- counts %>% filter(rowname %in% significant$ensembl_gene_id_version)
# counts <- left_join(counts, conversions, by=c("rowname" = "ensembl_gene_id_version"))
# 
# na_idx <- which(is.na(counts$external_gene_name))
# counts$external_gene_name[na_idx] <- counts$rowname[na_idx]
# em_idx <- (which(counts$external_gene_name == ""))
# counts$external_gene_name[em_idx] <- counts$rowname[em_idx]
# 
# row.names(counts) <- counts$external_gene_name
# counts <- counts %>% dplyr::select(hsc18, hsc20, hsc30, hsc35, hsc36, hsc3, hsc4, hsc23, hsc25, hsc33)
# write.csv(counts, "./significant_counts_edgeR.csv")
# 
# # Explore Minimums
# minsC <- apply(counts[1:5], 1, function(x) {min(x)})
# minsI <- apply(counts[6:10], 1, function(x) {min(x)})
# minsdf <- data.frame(control = minsC, infected = minsI)
# maxofmins <- apply(minsdf, 1, function(x) {max(x)})
# # min(maxofmins) == 12
# # So, there are counts in either Infected or Control for each Gene 
# 
# ### Averages
# avgC <- apply(counts[1:5], 1, function(x) {mean(x)})
# avgI <- apply(counts[6:10], 1, function(x) {mean(x)})
# avgdf <- data.frame(control = avgC, infected = avgI)
# write.csv(avgdf, "/Shuai_RNA/ANALYSIS/LimmaEdgeR/sig_AVGcounts_edgeR.csv")
# 
# avgC <- apply(cpms[1:5], 1, function(x) {mean(x)})
# avgI <- apply(cpms[6:10], 1, function(x) {mean(x)})
# avgdf <- data.frame(control = avgC, infected = avgI)
# write.csv(avgdf, "/Shuai_RNA/ANALYSIS/LimmaEdgeR/sig_AVGcpms_edgeR.csv")
# 
# lcpms <- log2(cpms)
# avgC <- apply(lcpms[1:5], 1, function(x) {mean(x)})
# avgI <- apply(lcpms[6:10], 1, function(x) {mean(x)})
# avgdf <- data.frame(control = avgC, infected = avgI)
# write.csv(avgdf, "/Shuai_RNA/ANALYSIS/LimmaEdgeR/sig_AVGlog2cpms_edgeR.csv")
```

**No Genes were found to be significant**